Concurrent Programming in Harmony

### 2021-07-25 book/paper.tex
Latest

Harmony is heavily influenced by Leslie Lamport's work on
TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09}, as well as
Gerard Holzmann's work on Promela and SPIN~\cite{SPIN}.
Harmony is designed to have a lower learning curve than those
systems, but is not as powerful.  When you finish this book
and want to learn more, we strongly encourage checking
those out.
Another excellent resource is Fred Schneider's book ``On
Concurrent Programming''~\cite{Schneider97}.
(This chapter is named after that book.)

@@

\section*{Exercises}
\begin{problems}
\item Check if the code finds a unique leader if identifiers are not unique.
\end{problems}

@@

The statement $x~+$$= 3$ loads $x$, adds 3, and then stores the
results in $x$.  In this case, it is equivalent to $x = x + 3$.
However, in general this is not so.  For example, $x[\mathtt{f}()]~+$$= 3$
only evaluates $\mathtt{f}()$ once.  Unlike Python, however,
$x~+$$= [3]$ is equivalent to $x = x + [3]$ in Harmony.
(In Python, the following two compound statements lead to different results
for $y$: $x = y = []; x~+$$= [1]$ and $x = y = []; x = x + [1]$.)

@@

The expression \textbf{const} \texttt{N = 3} introduces a new constant
\texttt{N} with the value 3.  Evaluating \texttt{N} does not lead to
loading from a memory location.  The assignment can be overridden with
the \texttt{-c} flag: \texttt{harmony -cN=4} executes the model checker
with 4 assigned to \texttt{N} instead of 3.
Harmony also supports \textbf{const} \texttt{N, M = 3, 4}, which
assigns 3 to \texttt{N} and 4 to \texttt{M}.
Harmony also has limited support for \emph{constant folding}.  For example,
\textbf{const} \texttt{N = 3 + 4} assigns value 7 to constant \texttt{N}.

@@

The statement \textbf{for} $x$ \textbf{in} $y: S_1; S_2; ...$
iterates over $y$ and executes for each element the statements
$S_1, S_2, ...$.
which must be a set or a dictionary.  $y$ is evaluated only once at
the beginning of the evaluation of this statement.  In case of a set,
the result is sorted (using Harmony's global order on all values).
For each element, the statements $S_1, S_2, ...$ are executed with
local variable $y$ having the value of the element.
$x$ can be a pattern such as $(a)$ or $(a, (b, c))$.
If the pattern cannot be matched, Harmony detects and error.
It is allowed, but discouraged, to assign different values to $x$
within statements $S_1, S2, ...$.

@@

You can import all constants from a module \texttt{m}
(including program counter constants) using the statement
\textbf{from} \texttt{m} \textbf{import} \texttt{*}.
This, however, excludes constants that start with the character \_: those
are considered \emph{private} to the module.

@@

The statement \textbf{go} $c e$ starts a thread with context $c$ that
has executed a \textbf{stop} expression.  The \textbf{stop} expression
returns value $e$.
The same context can be started multiple times, allowing threads to
\emph{fork}.

@@

The statement \textbf{invariant} $c$ declares that boolean expression $c$
is an invariant.  $c$ is only allowed to read shared variables and
is evaluated atomically after every state change.
If it evaluated to \texttt{False} Harmony reports and error.  Harmony
also reports an error if the expression evaluates to a value other
than \texttt{False} or \texttt{True}.

@@

\begin{code}
\begin{verbatim}
x = y = Lock();
\end{verbatim}
\end{code}

@@

Each (non-reserved) identifier in a Harmony program refers to either
a constant, a global variable, or a local variable.
Constants are declared using \texttt{const} statements.
Those constants are computed at compile-time.

Local variables all declared.  They can be declared in
\texttt{def} statements (i.e., arguments),
\texttt{let} statements, and in \texttt{for} loops.
Also, each method has an implicitly declared
\textit{result} variable, which is initialized to \texttt{None}.
Local variables are tightly scoped and cannot be shared
between threads.
While in theory one method can be declared within another,
they cannot share local variables either.
All other variables are global and must be initialized
before any threads are spawned.

@@

\section{For Loops and Comprehensions}

@@

Harmony also supports set, list, and dictionary comprehensions.
\index{comprehensions}%
Comprehensions are similar to Python, except that filtering uses
the keyword \texttt{where} instead of \texttt{if}.

>>>>

Harmony is heavily influenced by Leslie Lamport's work on
TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09},
Gerard Holzmann's work on Promela and SPIN~\cite{SPIN},
and University of Washington's DSLabs system~\cite{MWA19}.
Harmony is designed to have a lower learning curve than those
systems, but is not as powerful.  When you finish this book
and want to learn more, we strongly encourage checking
those out.
Another excellent resource is Fred Schneider's book ``On
Concurrent Programming''~\cite{Schneider97}.
(This chapter is named after that book.)

@@

\section*{Exercises}
\begin{problems}
\item Check if the code finds a unique leader if identifiers are not unique.
\item Messages are added atomically to the network.  Is this necessary?
What happens if you remove the \textbf{atomic} keyword?  Explain what happens.
\end{problems}

@@

The statement $x~+$$= 3$ loads $x$, adds 3, and then stores the
results in $x$.  In this case, it is equivalent to $x = x + 3$.
However, in general this is not so.  For example, $x[\mathtt{f}()]~+$$= 3$
only evaluates $\mathtt{f}()$ once.  Unlike Python, however,
$x~+$$= [3,]$ is equivalent to $x = x + [3,]$ in Harmony.
(In Python, the following two compound statements lead to different results
for $y$: $x = y = []; x~+$$= [3]$ and $x = y = []; x = x + [3]$.)

@@

The expression \textbf{const} \texttt{N = 3} introduces a new constant
\texttt{N} with the value 3.  Evaluating \texttt{N} does not lead to
loading from a memory location.  The assignment can be overridden with
the \texttt{-c} flag: \texttt{harmony -cN=4} executes the model checker
with 4 assigned to \texttt{N} instead of 3.
Harmony also supports \textbf{const} \texttt{N, M = 3, 4}, which
assigns 3 to \texttt{N} and 4 to \texttt{M}.
Harmony has limited support for \emph{constant folding}.  For example,
\textbf{const} \texttt{N = 3 + 4} assigns value 7 to constant \texttt{N}.

@@

The statement \textbf{for} $x$ \textbf{in} $y: S_1; S_2; ...$
iterates over $y$ and executes for each element the statements
$S_1, S_2, ...$.
$y$ must be a set or a dictionary.  $y$ is evaluated only once at
the beginning of the evaluation of this statement.  In case of a set,
the result is sorted (using Harmony's global order on all values).
In case of a dictionary, the statement iterates over the values of
the dictionary in the order of the keys.  This makes iterating over
lists intuitive and identical to Python.
For each element, the statements $S_1, S_2, ...$ are executed with
local variable $y$ having the value of the element.
$x$ can be a pattern such as $(a)$ or $(a, (b, c))$.
If the pattern cannot be matched, Harmony detects and error.
It is allowed, but discouraged, to assign different values to $x$
within statements $S_1, S2, ...$.

If $y$ is a dictionary, Harmony also supports the form
\textbf{for} $k:v$ \textbf{in} $y: S_1; S_2; ...$.  This works similar,
except that $k$ is bound to the key and $v$ is bound to the value.

@@

You can import all constants from a module \texttt{m}
(including program counter constants) using the statement
\textbf{from} \texttt{m} \textbf{import} \texttt{*}.
This, however, excludes constants whose names start with the character
\texttt{\_}: those are considered \emph{private} to the module.

@@

The statement \textbf{go} $c$ $e$ starts a thread with context $c$ that
has executed a \textbf{stop} expression.  The \textbf{stop} expression
returns value $e$.
The same context can be started multiple times, allowing threads to
\emph{fork}.

@@

The statement \textbf{invariant} $c$ declares that boolean expression $c$
is an invariant.  $c$ is only allowed to read shared variables and
is evaluated atomically after every state change.
If it ever evaluates to \texttt{False} Harmony reports and error.  Harmony
also reports an error if the expression evaluates to a value other
than \texttt{False} or \texttt{True}.

@@

\begin{code}
\begin{verbatim}
x = y = Lock()
\end{verbatim}
\end{code}

@@

Each (non-reserved) identifier in a Harmony program refers to either
a constant, a global shared variable, a local variable, or a module.
Constants are declared using \texttt{const} statements.
Those constants are computed at compile-time.

Local variables all declared.  They can be declared in
\texttt{def} statements (i.e., arguments),
\texttt{let} statements, \texttt{for} loops, and \texttt{select} statements.
Also, each method has an implicitly declared
\textit{result} variable, which is initialized to \texttt{None}.
Each thread has a variable called \textit{this} that contains
the thread-local state.
Local variables are tightly scoped and cannot be shared
between threads.
While in theory one method can be declared within another,
they cannot share local variables either.
All other variables are global and must be initialized
before any threads are spawned.

@@

\begin{comment}
\section{For Loops and Comprehensions}

@@

Harmony also supports set, list, and dictionary comprehensions.
\index{comprehensions}%
Comprehensions are similar to Python, except that filtering uses
the keyword \texttt{where} instead of \texttt{if}.
\end{comment}
